---
title: "Problem3B"
author: "Martine Middelthon and Helene Behrens"
date: "2 april 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem B
We consider data of bilirubin concentration in blood of three different patients. 
```{r}
bilirubin <- read.table("https://www.math.ntnu.no/emner/TMA4300/2020v/Exercise/ex3-additionalFiles/bilirubin.txt", header=T)
```
We plot this data for the three different individuals to compare their bilirubin concentrations. The resulting box-plots are seen below. 
```{r}
#  boxplot of logarithms of concentration for each indivicual:
boxplot(log(meas)~pers, data=bilirubin,
        main="Boxplot of Logarithm of Concentration", xlab="Individual", ylab="log(meas)")
```
From this plot, we observe a clear tendency that the bilirubin levels are higher for person 3 (p3) than for person 1 and person 2. We want to invesigate wether this difference is significant. We use the $\texttt{lm}$ function to fit the regression model 
$$
\log(Y_{ij}) = \beta_i + \epsilon_{ij}, i=1,2,3, j=1,\ldots,n_i;
$$
where $n_1=11, n_2=10$ and $n_3=8$, and $\epsilon \sim N(0, \sigma^2).$ The results are shown below. 
```{r}
reg <- lm(log(meas)~pers, data=bilirubin)
summary.lm <- summary(reg); summary.lm
Fval <- summary.lm$fstatistic; Fval
```
We want test the hypothesis that $\beta_1=\beta_2=\beta_3,$ i.e. that the expected bilirubin levels are actually the same for all individuals. To check this, we use the F-test with the reported F statistic from the $\texttt{summary}$, `{r}Fval`. The reported p-value of this F-statistic is 0.039, and thus we can reject the hypothesis that $\beta_1=\beta_2=\beta_3$ at significance level 5$\%$. Alternatively, we can argue that the 95$\%$ quantile of the F-distribution with 2 and 26 degrees of freedom is `qf(.95, df1=2, df2=26)`, which is less than $\texttt{Fval}$, which gives the same conclusion. 
```{r}
#  defining function PermTest():
PermTest <- function(bilirubin){
  perm.df <- data.frame(meas=sample(bilirubin$meas), pers = bilirubin$pers)
  reg <- lm(log(meas)~pers, data=perm.df)
  Fval <- summary(reg)$fstatistic["value"]
  return(Fval)
}
```
We perform a perturbation test between the three individuals to check whether the value of the F-statistic is due to an actual difference in the bilirubin level in the individuals, or by chance in the distribution of the measurements between the individuals. Using the $\texttt{PermTest}$-function defined above, we generate 999 samples of the F-statistic:
```{r}
F.samp <- rep(0,999)
for(i in 1:999){
  F.samp[i] <- PermTest(bilirubin)
}
```
From these, we compute the p-value as the probability to observe an F-statistic as extreme or more extreme than our original F-statistic, $\texttt{Fval}$, if there is no difference in the bilirubin levels of the individuals (i.e. if one can change the distribution of measurements between the individuals.)
```{r}
#  compute p-value for Fval using F.samp:
p.val <- length(F.samp[F.samp>=Fval["value"]])/999; p.val
```
The p-value is 0.044, so at a significance level 5$\%$ we reject the hypothesis that the bilirubin levels are the same for all individuals. 

